---
title: "Reproducible Research: Peer Assessment 1"
author: "Hidde Boom"
date: "Friday, February 19, 2021"
output: html_document
keep_md: true
---


## Loading and preprocessing the data
Unzip the data file and read the activity csv file.

```{r}
unzip("./activity.zip")
activityData <- read.csv("./activity.csv")
```

### Explore the dataset
The variables in the data are:

```{r}
names(activityData)
```

A summary of the data:
```{r}
summary(activityData)
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day:
```{r}
stepsPerDay <- aggregate(steps ~ date, activityData, sum, na.rm = T)
head(stepsPerDay) 
```

2. Make a histogram of the total number of steps taken each day
```{r}
hist(stepsPerDay$steps, col = "steelblue", xlab = "Steps", main = "Histogram of steps per day and the frequency")
```

3. Calculate and report the mean and median of the total number of steps taken per day
Mean:
```{r}
(meansStepsPerDay <- mean(stepsPerDay$steps))
```
Median:
```{r}
(medianStepsPerDay <- median(stepsPerDay$steps))
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
stepsPerInterval <- aggregate(steps ~ interval, activityData, mean, na.rm = T)
plot(steps ~ interval, stepsPerInterval, type = "l", xlab = "5-minute interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
(MaxIntervalSteps <- stepsPerInterval[which.max(stepsPerInterval$steps),]$interval)
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs)
```{r}
(TotMissingValues <- sum(is.na(activityData$steps)))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

* Chosen strategy is to use the mean per interval to fill the missing values
```{r}
MeanStepsPerInterval <- function(interval){
      stepsPerInterval[stepsPerInterval$interval == interval,]$steps
}
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activityDataWithoutNA <- activityData
for(i in 1:ncol(activityDataWithoutNA)){
      if(is.na(activityDataWithoutNA[i,]$steps)){
            activityDataWithoutNA[i,]$steps <- MeanStepsPerInterval(activityDataWithoutNA[i,]$interval)
      }
}
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
stepsPerDayWithoutNA <- aggregate(steps ~ date, activityDataWithoutNA, sum)

hist(stepsPerDayWithoutNA$steps, col = "darkred", xlab = "Steps", main = "Histogram of steps per day and the frequency, without NA")
```

Mean:
```{r}
(meansStepsPerDayWithoutNA <- mean(stepsPerDayWithoutNA$steps))
```
Median:
```{r}
(medianStepsPerDayWithoutNA <- median(stepsPerDayWithoutNA$steps))
```

comparison of mean and median between two datasets:

x            | With NA   | Without NA
------------|-----------|------------
Mean        | 10766.2   | 10566.9
Median      | 10765.0   | 10682.5


## Are there differences in activity patterns between weekdays and weekends?

For this part the \color{red}{\verb|weekdays()|}weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
activityDataWithoutNA$date <- as.Date(strptime(activityDataWithoutNA$date, format = "%Y-%m-%d"))
activityDataWithoutNA$day <- weekdays(activityDataWithoutNA$date)

for(i in 1:nrow(activityDataWithoutNA)) {
      if (activityDataWithoutNA[i, ]$day %in% c("Saturday", "Sunday")) {
            activityDataWithoutNA[i, ]$day <- "weekend"
      }
      else {
            activityDataWithoutNA[i, ]$day <- "weekday"
      }
}
```

2. Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r}
StepsByDay <- aggregate(steps ~ interval + day, activityDataWithoutNA, mean)

library("lattice")
lattice.options(
      layout.heights=list(bottom.padding=list(x=0), top.padding=list(x=0)),
      layout.widths=list(left.padding=list(x=4), right.padding=list(x=4))
)
xyplot(steps ~ interval | day, StepsByDay, type = "l", layout = c(1,2), 
       xlab = "5-minute Interval",
       ylab = "Number of steps", 
       margin = T)
```

## **NOTE**: for some reason the upper plot (weekend) does not appear, running the code normally gives both figures though. The figure is attached to the Github repo (named *PanelPlot*).







